---
title: "depth-analysis"
author: "Will oestreich"
date: '2024-10-02'
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, # show warnings
  message = FALSE, # show messages
  error = TRUE, # do not interrupt generation in case of errors,
  echo = TRUE  # show R code
)
```


```{r}
library(tidyverse) 
library(dplyr)
library(sf) 
library(terra)
library(raster)
library(arrow)
library(stringr)
library(MetBrewer)
```

## load in parquet
```{r}
obis_snapshot <- "obis_20240625.parquet"
```

## open dataset
```{r}
db <- open_dataset(obis_snapshot)
```

## quality filters
```{r}
obis_filter <- db |>
  filter(!str_detect(flags, "DATE_IN_FUTURE")) |> 
  filter(!str_detect(flags, "DATE_BEFORE_MIN")) |> 
  filter(!str_detect(flags, "NO_COORD")) |> 
  filter(!str_detect(flags, "ZERO_COORD")) |> 
  filter(!str_detect(flags, "LON_OUT_OF_RANGE")) |> 
  filter(!str_detect(flags, "LAT_OUT_OF_RANGE")) |> 
  filter(!str_detect(flags, "NO_DEPTH")) |> 
  filter(!str_detect(flags, "ON_LAND")) |> 
  mutate(bath_obs_prop_min = minimumDepthInMeters/bathymetry) |>
  mutate(bath_obs_prop_max = maximumDepthInMeters/bathymetry) |>
  mutate(bathymetry = round(as.numeric(bathymetry), 0)) |>
  mutate(minimumDepthInMeters = round(as.numeric(minimumDepthInMeters), 0)) |>
  mutate(depth_diff = bathymetry - minimumDepthInMeters) |>
  mutate(depth_quality_category = ifelse(bath_obs_prop_min <= 1.05 & bath_obs_prop_min >= 0, "depth_good", "depth_bad")) |>
  mutate(latitude = round(decimalLatitude,3), 
         longitude = round(decimalLongitude,3)) |>
  group_by(bathymetry, minimumDepthInMeters, bath_obs_prop_min, bath_obs_prop_max, depth_quality_category, latitude, longitude) |>
  count() |> 
  collect()
```

## explore outcome of custom depth quality flag
```{r}
depth_category_sums <- obis_filter |>
  group_by(depth_quality_category) |>
  summarize(total_n = sum(n))
head(depth_category_sums)
```

## filter out records with bad depth vs. bathymetry comparison flag
```{r}
obis_clean <- obis_filter |> 
  filter(depth_quality_category == "depth_good")
```

## the custom flag lets in records with depth up to 5% greater than bottom depth. for these records, assign depth to bottom depth and make the proportional depth = 1
```{r}
obis_clean <- obis_clean |>
  mutate(bath_obs_prop_min = ifelse(bath_obs_prop_min > 1, 1, bath_obs_prop_min)) |>
  mutate(minimumDepthInMeters = ifelse(bath_obs_prop_min > 1, bathymetry, minimumDepthInMeters))
```

## plot a weighted histogram of observation depth 
```{r}
obis_clean |> 
  ggplot(aes(x = minimumDepthInMeters, weight = n)) +
  geom_histogram(binwidth = 100) +
  labs(x = "min_depth", y = "Weighted Count") +
  theme_minimal()
```

## plot a weighted histogram of observation depth proportional to bottom depth
```{r}
obis_clean |> 
  ggplot(aes(x = bath_obs_prop_min, weight = n)) +
  geom_histogram(binwidth = 0.02) +
  labs(x = "min_depth prop", y = "Weighted Count") +
  theme_minimal()
```

## zones
```{r}
obis_clean <- obis_clean %>%
  mutate(zone = case_when(
    bathymetry < 200 ~ "coastal_or_shallow",
    bathymetry >= 200 & minimumDepthInMeters < 200 ~ "coastal_or_shallow",
    bathymetry >= 200 & minimumDepthInMeters > 200 & bath_obs_prop_min < 0.99 ~ "deep",
    bathymetry >= 200 & minimumDepthInMeters > 200 & bath_obs_prop_min >= 0.99 ~ "benthic"
  ))

obis_zone_summary <- obis_clean %>%
  group_by(zone) %>%
  summarise(total_n = sum(n, na.rm = TRUE))
obis_zone_summary$pct <- obis_zone_summary$total_n/sum(obis_zone_summary$total_n, na.rm = TRUE)
head(obis_zone_summary)
```

## maps
```{r}
df_spatial <- obis_clean |> 
  dplyr::filter(!is.na(latitude), 
         !is.na(longitude)) |> 
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m")
```

```{r}
zone_list <- split(df_spatial, df_spatial$zone)

raster_list <- list()

crs <- "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m" 
for (zone_name in names(zone_list)) {
  zone_data <- zone_list[[zone_name]]
  
  ras_temp <- raster(xmn=-180, xmx=180, ymn=-90, ymx=90, resolution=c(0.1, 0.1), vals=NA)
  raster_zone <- rasterize(zone_data, ras_temp, field = "n", fun = 'sum')
  
  terra_zone <- terra::rast(raster_zone)
  terra_zone_proj <- terra::project(terra_zone, crs, mask=TRUE)
  
  raster_list[[zone_name]] <- terra_zone_proj
}

raster_stack <- terra::rast(raster_list)
```

```{r global_plot, cache=TRUE}
colors <- c("grey", met.brewer(name="Isfahan1",n=20,type="continuous"))
# take log for viz
terra::plot(log(raster_stack,10), col = colors)
```
